The Precambrian paleogeography of Laurentia

### 2020-10-02 Manuscript/Laurentia_Paleogeo_Manuscript.tex
Baltica connection edits

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogeny of Baltica is a continuation of the the Mazatzal-Labradorian orogenies of Laurentia\citep{Karlstrom2001a}. Accretionary orogenesis within the Gothian orogeny was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from 
the Salla dike is considered to be representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic. This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b}. The timing of the main phase of the Sveconorwegian orogeny from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. However, it is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga.

>>>>

Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland \citep{Gower1990a, Salminen2021b}. In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin wherein the Gothian orogen of Baltica, where accretionary orogenesis was active ca. 1.66 to 1.52 Ga \citep{Bergstrom2020a}, is a continuation of the Mazatzal-Labradorian orogenic belts of Laurentia\citep{Karlstrom2001a}. Baltica and Laurentia as conjoined cratons with a shared active margin features as a major component of Paleoproterozoic to Mesoproterozoic paleogeographic reconstructions \citep{Evans2011a, Zhang2012a, Elming2021a}. Rotating Baltica into the NENA connection position results in matched paleomagnetic pole pairs between ca. 1.75 to 1.26 Ga \citep{Buchan2000a, Evans2008a} that could be extended to ca. 1.12 Ga if a virtual geomagnetic pole from the Salla dike is considered to be a representative paleomagnetic pole for Baltica \citep{Salminen2009b}. These data constrain the NENA connection to have been maintained until at least 1.26 Ga and perhaps until 1.12 Ga \citep{Salminen2021b}.

Many paleogeographic models consider there to still have been close proximity between Baltica and Laurentia in the latest Mesoproterozoic into the Neoproterozoic. This continued connection has been hypothesized based on correlation of the Grenvillian orogeny with the Sveconorwegian orogeny of southwestern Baltica\citep{Gower1990b} as well as proposed similarities between the apparent polar wander path of Laurentia's Grenville loop and Baltica's Sveconorwegian loop that led reconstructions to be based on the alignment of these paths and the resulting position of Baltica relative to Laurentia \citep{Piper1980a,Pisarevsky2003a}. However, as geochronology has improved the ages of poles in the Sveconorwegian loop are now constrained to be younger than the Grenville loop poles falling in a gap in Laurentia's pole record \citep{Evans2015a,Fairchild2017a} which makes it more challenging to test proposed post-NENA configurations between the cratons.

In terms of the orogenic timing, the main phase of the Sveconorwegian orogeny was from ca. 1.05 to 0.98 Ga corresponds temporally with the Grenvillian orogeny \citep{Stephens2020a}. However, it is currently an active debate in the literature whether the Sveconorwegian orogen is the result of collisional or accretionary orogenesis \citep{Stephens2020a}. \cite{Slagstad2019a} favor an accretionary orogeny and by contrasting this setting with the collisional orogenesis of the Grenvillian orogen argue that there should not be a link between Laurentia and Baltica ca. 1.0 Ga. However, in a recent critical review of the constraints, the high pressure nature of metamorphism and the cratonward propagation of orogenesis (in contrast to older accretionary orogenesis) led \cite{Stephens2020a} to favor the interpretation that the Sveconorwegian orogen is a record of prolonged continent-continent collision initiating ca. 1.06 Ga.

Another basis for a continued connection between
